Flute and kink instabilities in a dynamically twisted flux tube with anisotropic plasma viscosity

### 2021-12-02 paper.tex
Change fluting to flute everywhere

In general, the stability of a cylindrical twisted magnetic flux tube is analysed using perturbations of the form
\begin{equation}
  \label{eq:kink_perturbation}
\xi(r, \theta, z) = \xi(r) e^{i(m\theta + kz)},
\end{equation}
where $m$ and $k$ are the wavenumbers \rs{in the $\theta$ and $z$
  directions, respectively}{in the azimuthal and axial directions,
  $\theta$ and $z$, respectively and $r$ is the radial coordinate in
  cylindrical polars}. 
The helical kink instability occurs for perturbations where $m=1, k\ne0$ and is the only instability of this form which is a body instability, that is it moves the entire body of the flux tube. Perturbations where $m>1$ are termed \rs{fluting}{flute} or interchange instabilities.

When the magnetic field is sheared, as in a twisted magnetic flux
tube, an interchange instability (such as the \rs{fluting}{flute}
instability) is confined to a surface where the peaks and troughs
follow the shear of the field. That is, the instability is confined to
the surface where the perturbation wavevector $(0, m/r, k)$ is
perpendicular to the direction of the field, \rs{}{known as the
  ``resonance surface''}. \rs{which in}{In} an axisymmetric twisted
flux tube \rs{}{the resonance surface} is located at a radius $r$
given by
\begin{equation}
  \label{eq:resonant_surface}
\frac{m}{r} B_{\theta}(r) + kB_z(r) \approx 0.
\end{equation}

The stability of an infinite cylindrical flux tube to perturbations of
the form~\eqref{eq:kink_perturbation} is given by the classical
\rs{}{Suydam's criterion~\citep{suydamStabilityLinearPinch1958}}
\begin{equation}
  \label{eq:suydams_criterion}
\frac{B_z^2 S^2}{4} + 2 r p' > 0,
\end{equation}
where $S = r q'/q$ is a measure of the shear, $q = 2\pi r B_z / L
B_{\theta}$ is the safety factor for a flux tube of length $L$ and a
\rs{dash}{prime} denotes differentiation with respect to
$r$~\citep{mikhailovskiiInstabilitiesConfinedPlasma1998}. This applies
to both \rs{fluting}{flute} and kink instabilities although many
additional effects such as line-tying are not incorporated into the
corresponding linear analysis. The effect of line-tying on the kink
instability \rs{can be found}{is investigated}
in~\citet{hoodKinkInstabilitySolar1979}. Where~\eqref{eq:suydams_criterion}
is not satisfied, the flux tube may be unstable to perturbations of
the form~\eqref{eq:kink_perturbation}. When $m>1$, the perturbations
remain local to resonant surfaces given
by~\eqref{eq:resonant_surface}. 
\rs{}{When} Suydam's criterion is satisfied and the flux tube is linearly stable, it may still be unstable to non-local perturbations, where the shear and pressure are small enough that interchange perturbations do not need to satisfy~\eqref{eq:resonant_surface}. Additionally, the inclusion of resistivity generally reduces the stabilising effect of the shear, permitting growth of a resistive interchange mode, albeit at a slower rate than that of the ideal instability~\citep{mikhailovskiiInstabilitiesConfinedPlasma1998}. It will be found that the ideal linear analysis of~\cite{mikhailovskiiInstabilitiesConfinedPlasma1998} is sufficient for understanding the \rs{fluting}{flute} instabilities investigated here since the associated flux tubes adequately fail the criterion~\eqref{eq:suydams_criterion}.

While Suydam's condition gives an indication of the stability of a flux tube to a given perturbation, the linear growth rate of the ideal \rs{fluting}{flute} instability $\gamma$ can be found via a stability analysis analogous to that of the Rayleigh-Taylor instability (see~\citep[][]{goldstonIntroductionPlasmaPhysics2020}) and is given by
\begin{equation}
  \label{eq:fluting_growth_rate}
\gamma^2 = \frac{2|\nabla p|}{\rho R_c},
\end{equation}
where $R_c$ is the radius of curvature of the magnetic field. This equation only applies when the pressure gradient and radius of curvature vector are in the same direction, that is the plasma is constrained by a concave magnetic field such that the pressure forces and magnetic tension forces are in competition. In a cylindrical, twisted flux tube, the field is always concave towards the central axis of the tube, so any inwardly directed pressure gradient is potentially unstable to fluting.

@@

The flux rope is formed by prescribing a slowly accelerating, rotating flow at the upper $z$-boundary as
\begin{equation}
  \label{eq:null_twisting_profile}
  \vec{u} = u_0 u_r(r) u_t(t) (-y, x, 0)^T,
\end{equation}
where $u_r(r)$ describes the radial profile of the twisting motion in terms of the radius $r^2 = x^2 + y^2$,
\begin{equation}
  \label{eq:radial_twisting_function}
  u_r(r) = u_{r0}(1 + \tanh(1 - r_d r^2)),
\end{equation}
where $r_d$ controls the radial extent of the driver, $u_{r0}$ is a normalising factor, and $u_t(t)$ describes the imposed acceleration of the twisting motion,
\begin{equation}
  \label{eq:ramping_up_function}
  u_t(t) = \tanh^2(t/t_r),
\end{equation}
where the parameter $t_r$ controls the time taken to reach the final driver velocity $u_0$. The functions $u_r(r)$ and $u_t(t)$ are plotted in figure~\ref{fig:kink_driver}. At the lower boundary, the flow is in the opposite direction. This form of driver allows the system to be accelerated slowly enough that the production of disruptive shocks and fast waves is minimal. It is unavoidable that some waves are produced during the boundary acceleration, however these usefully provide \rs{one}{a} source of noise which eventually forms a perturbation.
\rs{}{
The driver} velocity is set to $u_0 = 0.15$, the normalising factor is $u_{r0} = 2.08$, and setting $r_d = 5$ corresponds to a driver constrained to $r<1$ and with a peak velocity at $r\approx 0.38$. The ramping time is set to $t_r = 2$, resulting in an acceleration from $0$ to $u_0$ over approximately $5$ Alfv\'en times. These driver parameters correspond to a peak rotational period of $T_R = 15.92$, the length of time taken for one full turn to be injected by a single driver. Both drivers result in twist being added at a rate of $2\pi$ every $7.96$ Alfv\'en times. The twist profile across the entire flux tube develops in such a way that by $t\approx 20$, it is qualitatively similar to those studied in~\citet{quinnEffectAnisotropicViscosity2020,hoodCoronalHeatingMagnetic2009,barefordShockHeatingNumerical2015} however the length of the flux tubes differs significantly. This configuration produces a $z$-directed tube of increasingly twisted magnetic field that eventually becomes unstable to both the \rs{fluting}{flute} instability and the helical kink instability.

@@

This current profile is reflected in the radial Ohmic heating profile
(figure~\ref{fig:kink_straight_ohmic_heating_profile}) and,
consequently, in the radial pressure profile
(figure~\ref{fig:pressure_profiles}). The highly pressurised core
extends to $r\approx 0.2$--$0.4$ (depending on the value of $\eta$)
before increasing slightly around $r\approx 0.7$. The secondary bump
in pressure is due to the outer shell of current. The pressure
gradient near the axis provides the outwardly directed pressure force
which competes against the binding action of the magnetic tension
\rs{to (potentially) result in the}{and this provides the mechanism of}
\rs{fluting}{flute} instability \rs{}{excitation}.  The magnitude
of the pressure gradient depends strongly on the value of
\rs{}{resistivity} $\eta$, with lower values producing shallower 
gradients which (as shall be seen) are more stable to the
\rs{fluting}{flute} instability. Indeed, when $\eta=0$, the radial
pressure profile is nearly flat and the tube stable to the
\rs{fluting}{flute} instability.

In all cases unstable to the \rs{fluting}{flute} instability, it
occurs \rs{some time}{} between $t=20$ and $t=30$. \rs{During this
  time, the}{The} continued driving at the boundaries eventually
injects enough twist that the tube also becomes unstable to the kink
instability. This initially develops linearly alongside \rs{}{or shortly
after} the \rs{fluting}{flute} instability and then erupts during the
\rs{kink's}{its}  nonlinear phase, dominating the dynamics and
disrupting the   \rs{fluting}{flute} instability. The \rs{}{onset and
the competition} of the two instabilities is strongly affected by
the value of $\eta$ and the viscosity model used.

@@

\begin{figure*}
  \centering
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{swi-4_pressure_13.pdf}
      \caption{$t=26$}
      \label{fig:swi-4_pressure_13}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{swi-4_pressure_14.pdf}
      \caption{$t=28$}
      \label{fig:swi-4_pressure_14}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{swi-4_pressure_15.pdf}
      \caption{$t=30$}
      \label{fig:swi-4_pressure_15}
    \end{subfigure}
\mycaption{Pressure profiles during the development of the
  \rs{fluting}{flute} and kink instabilities.}{Shown are
  \rs{slices}{density plots} of pressure \rs{through}{at} $z=0$
  \rs{where}{with} $\eta = 10^{-4}$ 
  and \rs{the viscosity model is \rs{switching}{anisotropic}}{for the
    anisotropic viscosity model}. Note the difference in colour scale
  in figure~\ref{fig:swi-4_pressure_15}. The development in the
  isotropic case is similar.} 
\label{fig:kink_pressure_slices-4}%
\end{figure*}

\rs{}{We now describe the evolution and competition of flute and
  kink instabilities in case of resistivity $\eta=10^{-4}$.}
\rs{The}{Figure}~\ref{fig:kink_pressure_slices-4} shows the pressure profile
\rs{the viscosity model is switching for times}{of the \rs{switching}{anisotropic} viscosity case 
\eqref{eq:switching_model} at
time moments} $t=26$, $28$ and $30$ and \rs{through}{at} $z=0$. At $t=26$
the tube becomes unstable to \rs{the $m=4$}{} \rs{fluting}{flute}
instability \rs{}{with azimuthal wavenumber $m=4$}, when the
plasma begins to \rs{(slightly)}{} bulge out diagonally from the
high-pressure core as seen in figure~\ref{fig:swi-4_pressure_13}. As  
the bulges move radially outwards into lower pressure regions they
expand and accelerate, resulting in the entire pressure structure
appearing \rs{clover-shaped}{taking the shape of a four-leaf clover} (figure~\ref{fig:swi-4_pressure_14}). By
$t=30$ the kink instability has disrupted the \rs{fluting}{flute}
instability and is developing nonlinearly
\rs{}{as evident in} figure~\ref{fig:swi-4_pressure_15}. As is typical of nonlinear kink
development, the tube continues to release its stored potential energy
\rs{as}{in the form of} kinetic energy and heat and the contained plasma becomes highly
mixed. In the \rs{unseen}{} isotropic \rs{}{viscosity} case
\rs{}{which will not be illustrated by a separate figure}, the
\rs{fluting}{flute} instability is present but damped, and \rs{}{it is
  quickly outcompeted by} the kink instability \rs{}{which} quickly dominates the dynamics.

\begin{figure}
  \centering
    \begin{subfigure}{0.49\textwidth}
      \includegraphics[width=\linewidth]{field_line_plots/cropped/v-4r-4-isotropic_0014_cropped.png}
      \caption{Isotropic}
      \label{fig:field_line_plots_iso}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.49\textwidth}
      \includegraphics[width=\linewidth]{field_line_plots/cropped/v-4r-4-switching_0014_cropped.png}
      \caption{\rs{Switching}{Anisotropic}}
      \label{fig:field_line_plots_swi}
    \end{subfigure}
\mycaption{Simultaneous development of \rs{fluting}{flute} and kink
  instabilities in the isotropic and \rs{switching}{anisotropic} cases
  \rs{as}{illustrated by} field lines and pressure contours.}{Field
  lines and contours of pressure (where $p=0.3$) are plotted at
  $t=28$. Also shown is the velocity driver \rs{as a slice}{$u_r(\sqrt{x^2+y^2})$ at $z=2$}. The \rs{fluting}{flute} instability grows in both cases, though faster in the \rs{switching}{anisotropic} case. The initial stages of the kink instability can also be observed in the field lines of the isotropic case in subfigure~\ref{fig:field_line_plots_iso}.}
\label{fig:kink_field_line_plots}%
\end{figure}

Figure~\ref{fig:kink_field_line_plots} shows the effect the viscosity
models have on the initial stages of the \rs{fluting}{flute} and kink
instabilities in 3D. While the \rs{fluting}{flute} instability is
observed in both cases, it is damped in the isotropic case and grows
faster in the \rs{switching}{anisotropic} case. In the latter case,
the extended development of the \rs{fluting}{flute} instability
appears to disrupt the inner core of field lines and, \rs{(as shall be
  seen)}{as will be discussed further below,} slows the growth of the
kink instability. In the isotropic case, the \rs{fluting}{flute}
instability has been damped to the extent that the kink instability
grows uninhibited and quickly disrupts the fluting.

\begin{figure}
  \centering
    \begin{subfigure}{0.49\textwidth}
      \includegraphics[width=\linewidth]{kinetic_energy-4.pdf}
      \caption{Kinetic Energy}
      \label{fig:kink_ke-4}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.49\textwidth}
      \includegraphics[width=\linewidth]{kinetic_energy_log-4.pdf}
      \caption{Growth rate estimation}
      \label{fig:kink_ke_log-4}
    \end{subfigure}
  \mycaption{Kinetic energy as a function of time showing the
    development and measured growth rates \rs{}{$\gamma$ and $\lambda$}
    of the \rs{fluting}{flute}
    and kink instabilities, \rs{}{respectively.}}{\rs{Both plots are from results
      where}{Resistivity is $\eta=10^{-4}$ and the second plot is a
      enlarged version of the first.}}
\label{fig:kink_str8_ke-4}%
\end{figure}

Despite the \rs{fluting}{flute} instability appearing in the isotropic
case (figure~\ref{fig:field_line_plots_iso}), only in the
\rs{switching}{anisotropic} case can the onset of both the
\rs{fluting}{flute} and kink instabilities be seen in the kinetic
energy profile \rs{}{shown in} figure~\ref{fig:kink_ke_log-4}. \rs{where}{Here,} the nonlinear
growth rates of the two instabilities are found to be $\gamma = 0.69$
for the \rs{fluting}{flute} and $\lambda = 2.55$ for the kink. The
onset times are approximately $t=27$ for the \rs{fluting}{flute}
instability and \rs{}{$t=29.5$} for
the kink. In the isotropic case, the growth rate of the kink, $\lambda
= 2.97$, is larger than in the \rs{switching}{anisotropic} case,
although the onset times appear similar, and the kinetic energy
profile shows no evidence of \rs{the growth of the \rs{fluting}{flute}
instability.}{\rs{fluting}{flute} instability growth.}

@@

\begin{figure}
  \centering
    \begin{subfigure}{0.49\textwidth}
      \includegraphics[width=\linewidth]{suydam_condition_4.pdf}
      \caption{Suydam condition}
      \label{fig:suydam_condition_4}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.49\textwidth}
      \includegraphics[width=\linewidth]{growth_rate_4.pdf}
      \caption{Linear growth rate}
      \label{fig:growth_rate_4}
    \end{subfigure}
\mycaption{Stability and linear growth rate of the \rs{fluting}{flute}
  instability.}{In \rs{}{panel} \ref{fig:suydam_condition_4}, Suydam's stability
  criterion \rs{}{\eqref{eq:suydams_criterion}} and is contributing
  terms \rs{(LHS of~\eqref{eq:suydams_criterion})}{}  are plotted and in
  \rs{}{panel} \ref{fig:growth_rate_4} the predicted linear growth rate~\eqref{eq:fluting_growth_rate2} is plotted. Both plots are produced at $t=20$ for $\eta=10^{-4}$ and using the \rs{switching}{anisotropic} model. The location of the peak linear growth rate is also shown.}
\label{fig:stability_and_growth}%
\end{figure}

@@

\begin{figure}
  \centering
    \begin{subfigure}{0.49\textwidth}
      \includegraphics[width=\linewidth]{perturbations_4.pdf}
      \caption{Perturbations}
      \label{fig:pressure_pert_4}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.49\textwidth}
      \includegraphics[width=\linewidth]{resonant_surface_4.pdf}
      \caption{Resonance function}
      \label{fig:resonant_surface_4}
    \end{subfigure}
\mycaption{Perturbations corresponding to the \rs{fluting}{flute} and
  kink instabilities and the spatial radial distribution of the
  associated resonance function.}{Pressure and velocity perturbations
  in $z$ (corresponding to the \rs{fluting}{flute} and kink
  instabilities, respectively) and of the resonance
  \rs{function}{surface} $m B_{\theta}(r)/r + kB_z(r)$ as a function
  of $r$ using the observed fluting perturbation wavenumbers. All
  plots are \rs{from}{} snapshots at $t=26$ where $\eta=10^{-4}$ and
  the viscosity model is \rs{switching}{anisotropic}.} 
\label{fig:k_and_resonance}%
\end{figure}

\rs{Figure~\ref{fig:pressure_pert_4} plots the }{The} observed
  perturbations corresponding to the \rs{fluting}{flute} and kink
  instabilities at $t=26$ \rs{}{are shown in
  Figure~\ref{fig:pressure_pert_4}}. The fluting perturbation is
  most easily observed in the pressure and is plotted as a function of $z$
  following a line through the point $(r, \theta) = (0.101, 0)$. The
  kink instability is best revealed in the $x$-velocity (a proxy for the
  radial velocity) through the axis. Comparing the magnitudes of the
  perturbations at this time suggests the \rs{fluting}{flute}
  instability is close to transitioning to its nonlinear phase while
  the kink instability is still very much in its linear
  phase.

@@

Comparing the effect of the viscous models on the perturbations, in the isotropic case, the fluting perturbation is damped, while in the \rs{switching}{anisotropic} case the kink perturbation is diminished, explaining why the \rs{fluting}{flute} instability appears more readily in the \rs{switching}{anisotropic} case (figure~\ref{fig:kink_ke-4}).

>>>>

In general, the stability of a cylindrical twisted magnetic flux tube is analysed using perturbations of the form
\begin{equation}
  \label{eq:kink_perturbation}
\xi(r, \theta, z) = \xi(r) e^{i(m\theta + kz)},
\end{equation}
where $m$ and $k$ are the wavenumbers \rs{in the $\theta$ and $z$
  directions, respectively}{in the azimuthal and axial directions,
  $\theta$ and $z$, respectively and $r$ is the radial coordinate in
  cylindrical polars}. 
The helical kink instability occurs for perturbations where $m=1, k\ne0$ and is the only instability of this form which is a body instability, that is it moves the entire body of the flux tube. Perturbations where $m>1$ are termed flute or interchange instabilities.

When the magnetic field is sheared, as in a twisted magnetic flux
tube, an interchange instability (such as the flute
instability) is confined to a surface where the peaks and troughs
follow the shear of the field. That is, the instability is confined to
the surface where the perturbation wavevector $(0, m/r, k)$ is
perpendicular to the direction of the field, \rs{}{known as the
  ``resonance surface''}. \rs{which in}{In} an axisymmetric twisted
flux tube \rs{}{the resonance surface} is located at a radius $r$
given by
\begin{equation}
  \label{eq:resonant_surface}
\frac{m}{r} B_{\theta}(r) + kB_z(r) \approx 0.
\end{equation}

The stability of an infinite cylindrical flux tube to perturbations of
the form~\eqref{eq:kink_perturbation} is given by the classical
\rs{}{Suydam's criterion~\citep{suydamStabilityLinearPinch1958}}
\begin{equation}
  \label{eq:suydams_criterion}
\frac{B_z^2 S^2}{4} + 2 r p' > 0,
\end{equation}
where $S = r q'/q$ is a measure of the shear, $q = 2\pi r B_z / L
B_{\theta}$ is the safety factor for a flux tube of length $L$ and a
\rs{dash}{prime} denotes differentiation with respect to
$r$~\citep{mikhailovskiiInstabilitiesConfinedPlasma1998}. This applies
to both flute and kink instabilities although many
additional effects such as line-tying are not incorporated into the
corresponding linear analysis. The effect of line-tying on the kink
instability \rs{can be found}{is investigated}
in~\citet{hoodKinkInstabilitySolar1979}. Where~\eqref{eq:suydams_criterion}
is not satisfied, the flux tube may be unstable to perturbations of
the form~\eqref{eq:kink_perturbation}. When $m>1$, the perturbations
remain local to resonant surfaces given
by~\eqref{eq:resonant_surface}. 
\rs{}{When} Suydam's criterion is satisfied and the flux tube is linearly stable, it may still be unstable to non-local perturbations, where the shear and pressure are small enough that interchange perturbations do not need to satisfy~\eqref{eq:resonant_surface}. Additionally, the inclusion of resistivity generally reduces the stabilising effect of the shear, permitting growth of a resistive interchange mode, albeit at a slower rate than that of the ideal instability~\citep{mikhailovskiiInstabilitiesConfinedPlasma1998}. It will be found that the ideal linear analysis of~\cite{mikhailovskiiInstabilitiesConfinedPlasma1998} is sufficient for understanding the flute instabilities investigated here since the associated flux tubes adequately fail the criterion~\eqref{eq:suydams_criterion}.

While Suydam's condition gives an indication of the stability of a flux tube to a given perturbation, the linear growth rate of the ideal flute instability $\gamma$ can be found via a stability analysis analogous to that of the Rayleigh-Taylor instability (see~\citep[][]{goldstonIntroductionPlasmaPhysics2020}) and is given by
\begin{equation}
  \label{eq:fluting_growth_rate}
\gamma^2 = \frac{2|\nabla p|}{\rho R_c},
\end{equation}
where $R_c$ is the radius of curvature of the magnetic field. This equation only applies when the pressure gradient and radius of curvature vector are in the same direction, that is the plasma is constrained by a concave magnetic field such that the pressure forces and magnetic tension forces are in competition. In a cylindrical, twisted flux tube, the field is always concave towards the central axis of the tube, so any inwardly directed pressure gradient is potentially unstable to fluting.

@@

The flux rope is formed by prescribing a slowly accelerating, rotating flow at the upper $z$-boundary as
\begin{equation}
  \label{eq:null_twisting_profile}
  \vec{u} = u_0 u_r(r) u_t(t) (-y, x, 0)^T,
\end{equation}
where $u_r(r)$ describes the radial profile of the twisting motion in terms of the radius $r^2 = x^2 + y^2$,
\begin{equation}
  \label{eq:radial_twisting_function}
  u_r(r) = u_{r0}(1 + \tanh(1 - r_d r^2)),
\end{equation}
where $r_d$ controls the radial extent of the driver, $u_{r0}$ is a normalising factor, and $u_t(t)$ describes the imposed acceleration of the twisting motion,
\begin{equation}
  \label{eq:ramping_up_function}
  u_t(t) = \tanh^2(t/t_r),
\end{equation}
where the parameter $t_r$ controls the time taken to reach the final driver velocity $u_0$. The functions $u_r(r)$ and $u_t(t)$ are plotted in figure~\ref{fig:kink_driver}. At the lower boundary, the flow is in the opposite direction. This form of driver allows the system to be accelerated slowly enough that the production of disruptive shocks and fast waves is minimal. It is unavoidable that some waves are produced during the boundary acceleration, however these usefully provide \rs{one}{a} source of noise which eventually forms a perturbation.
\rs{}{
The driver} velocity is set to $u_0 = 0.15$, the normalising factor is $u_{r0} = 2.08$, and setting $r_d = 5$ corresponds to a driver constrained to $r<1$ and with a peak velocity at $r\approx 0.38$. The ramping time is set to $t_r = 2$, resulting in an acceleration from $0$ to $u_0$ over approximately $5$ Alfv\'en times. These driver parameters correspond to a peak rotational period of $T_R = 15.92$, the length of time taken for one full turn to be injected by a single driver. Both drivers result in twist being added at a rate of $2\pi$ every $7.96$ Alfv\'en times. The twist profile across the entire flux tube develops in such a way that by $t\approx 20$, it is qualitatively similar to those studied in~\citet{quinnEffectAnisotropicViscosity2020,hoodCoronalHeatingMagnetic2009,barefordShockHeatingNumerical2015} however the length of the flux tubes differs significantly. This configuration produces a $z$-directed tube of increasingly twisted magnetic field that eventually becomes unstable to both the flute instability and the helical kink instability.

@@

This current profile is reflected in the radial Ohmic heating profile
(figure~\ref{fig:kink_straight_ohmic_heating_profile}) and,
consequently, in the radial pressure profile
(figure~\ref{fig:pressure_profiles}). The highly pressurised core
extends to $r\approx 0.2$--$0.4$ (depending on the value of $\eta$)
before increasing slightly around $r\approx 0.7$. The secondary bump
in pressure is due to the outer shell of current. The pressure
gradient near the axis provides the outwardly directed pressure force
which competes against the binding action of the magnetic tension
\rs{to (potentially) result in the}{and this provides the mechanism of}
flute instability \rs{}{excitation}.  The magnitude
of the pressure gradient depends strongly on the value of
\rs{}{resistivity} $\eta$, with lower values producing shallower 
gradients which (as shall be seen) are more stable to the
flute instability. Indeed, when $\eta=0$, the radial
pressure profile is nearly flat and the tube stable to the
flute instability.

In all cases unstable to the flute instability, it
occurs \rs{some time}{} between $t=20$ and $t=30$. \rs{During this
  time, the}{The} continued driving at the boundaries eventually
injects enough twist that the tube also becomes unstable to the kink
instability. This initially develops linearly alongside \rs{}{or shortly
after} the flute instability and then erupts during the
\rs{kink's}{its}  nonlinear phase, dominating the dynamics and
disrupting the   flute instability. The \rs{}{onset and
the competition} of the two instabilities is strongly affected by
the value of $\eta$ and the viscosity model used.

@@

\begin{figure*}
  \centering
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{swi-4_pressure_13.pdf}
      \caption{$t=26$}
      \label{fig:swi-4_pressure_13}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{swi-4_pressure_14.pdf}
      \caption{$t=28$}
      \label{fig:swi-4_pressure_14}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{swi-4_pressure_15.pdf}
      \caption{$t=30$}
      \label{fig:swi-4_pressure_15}
    \end{subfigure}
\mycaption{Pressure profiles during the development of the
  flute and kink instabilities.}{Shown are
  \rs{slices}{density plots} of pressure \rs{through}{at} $z=0$
  \rs{where}{with} $\eta = 10^{-4}$ 
  and \rs{the viscosity model is \rs{switching}{anisotropic}}{for the
    anisotropic viscosity model}. Note the difference in colour scale
  in figure~\ref{fig:swi-4_pressure_15}. The development in the
  isotropic case is similar.} 
\label{fig:kink_pressure_slices-4}%
\end{figure*}

\rs{}{We now describe the evolution and competition of flute and
  kink instabilities in case of resistivity $\eta=10^{-4}$.}
\rs{The}{Figure}~\ref{fig:kink_pressure_slices-4} shows the pressure profile
\rs{the viscosity model is switching for times}{of the \rs{switching}{anisotropic} viscosity case 
\eqref{eq:switching_model} at
time moments} $t=26$, $28$ and $30$ and \rs{through}{at} $z=0$. At $t=26$
the tube becomes unstable to \rs{the $m=4$}{} flute
instability \rs{}{with azimuthal wavenumber $m=4$}, when the
plasma begins to \rs{(slightly)}{} bulge out diagonally from the
high-pressure core as seen in figure~\ref{fig:swi-4_pressure_13}. As  
the bulges move radially outwards into lower pressure regions they
expand and accelerate, resulting in the entire pressure structure
appearing \rs{clover-shaped}{taking the shape of a four-leaf clover} (figure~\ref{fig:swi-4_pressure_14}). By
$t=30$ the kink instability has disrupted the flute
instability and is developing nonlinearly
\rs{}{as evident in} figure~\ref{fig:swi-4_pressure_15}. As is typical of nonlinear kink
development, the tube continues to release its stored potential energy
\rs{as}{in the form of} kinetic energy and heat and the contained plasma becomes highly
mixed. In the \rs{unseen}{} isotropic \rs{}{viscosity} case
\rs{}{which will not be illustrated by a separate figure}, the
flute instability is present but damped, and \rs{}{it is
  quickly outcompeted by} the kink instability \rs{}{which} quickly dominates the dynamics.

\begin{figure}
  \centering
    \begin{subfigure}{0.49\textwidth}
      \includegraphics[width=\linewidth]{field_line_plots/cropped/v-4r-4-isotropic_0014_cropped.png}
      \caption{Isotropic}
      \label{fig:field_line_plots_iso}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.49\textwidth}
      \includegraphics[width=\linewidth]{field_line_plots/cropped/v-4r-4-switching_0014_cropped.png}
      \caption{\rs{Switching}{Anisotropic}}
      \label{fig:field_line_plots_swi}
    \end{subfigure}
\mycaption{Simultaneous development of flute and kink
  instabilities in the isotropic and \rs{switching}{anisotropic} cases
  \rs{as}{illustrated by} field lines and pressure contours.}{Field
  lines and contours of pressure (where $p=0.3$) are plotted at
  $t=28$. Also shown is the velocity driver \rs{as a slice}{$u_r(\sqrt{x^2+y^2})$ at $z=2$}. The flute instability grows in both cases, though faster in the \rs{switching}{anisotropic} case. The initial stages of the kink instability can also be observed in the field lines of the isotropic case in subfigure~\ref{fig:field_line_plots_iso}.}
\label{fig:kink_field_line_plots}%
\end{figure}

Figure~\ref{fig:kink_field_line_plots} shows the effect the viscosity
models have on the initial stages of the flute and kink
instabilities in 3D. While the flute instability is
observed in both cases, it is damped in the isotropic case and grows
faster in the \rs{switching}{anisotropic} case. In the latter case,
the extended development of the flute instability
appears to disrupt the inner core of field lines and, \rs{(as shall be
  seen)}{as will be discussed further below,} slows the growth of the
kink instability. In the isotropic case, the flute
instability has been damped to the extent that the kink instability
grows uninhibited and quickly disrupts the fluting.

\begin{figure}
  \centering
    \begin{subfigure}{0.49\textwidth}
      \includegraphics[width=\linewidth]{kinetic_energy-4.pdf}
      \caption{Kinetic Energy}
      \label{fig:kink_ke-4}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.49\textwidth}
      \includegraphics[width=\linewidth]{kinetic_energy_log-4.pdf}
      \caption{Growth rate estimation}
      \label{fig:kink_ke_log-4}
    \end{subfigure}
  \mycaption{Kinetic energy as a function of time showing the
    development and measured growth rates \rs{}{$\gamma$ and $\lambda$}
    of the flute
    and kink instabilities, \rs{}{respectively.}}{\rs{Both plots are from results
      where}{Resistivity is $\eta=10^{-4}$ and the second plot is a
      enlarged version of the first.}}
\label{fig:kink_str8_ke-4}%
\end{figure}

Despite the flute instability appearing in the isotropic
case (figure~\ref{fig:field_line_plots_iso}), only in the
\rs{switching}{anisotropic} case can the onset of both the
flute and kink instabilities be seen in the kinetic
energy profile \rs{}{shown in} figure~\ref{fig:kink_ke_log-4}. \rs{where}{Here,} the nonlinear
growth rates of the two instabilities are found to be $\gamma = 0.69$
for the flute and $\lambda = 2.55$ for the kink. The
onset times are approximately $t=27$ for the flute
instability and \rs{}{$t=29.5$} for
the kink. In the isotropic case, the growth rate of the kink, $\lambda
= 2.97$, is larger than in the \rs{switching}{anisotropic} case,
although the onset times appear similar, and the kinetic energy
profile shows no evidence of \rs{the growth of the flute
instability.}{flute instability growth.}

@@

\begin{figure}
  \centering
    \begin{subfigure}{0.49\textwidth}
      \includegraphics[width=\linewidth]{suydam_condition_4.pdf}
      \caption{Suydam condition}
      \label{fig:suydam_condition_4}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.49\textwidth}
      \includegraphics[width=\linewidth]{growth_rate_4.pdf}
      \caption{Linear growth rate}
      \label{fig:growth_rate_4}
    \end{subfigure}
\mycaption{Stability and linear growth rate of the flute
  instability.}{In \rs{}{panel} \ref{fig:suydam_condition_4}, Suydam's stability
  criterion \rs{}{\eqref{eq:suydams_criterion}} and is contributing
  terms \rs{(LHS of~\eqref{eq:suydams_criterion})}{}  are plotted and in
  \rs{}{panel} \ref{fig:growth_rate_4} the predicted linear growth rate~\eqref{eq:fluting_growth_rate2} is plotted. Both plots are produced at $t=20$ for $\eta=10^{-4}$ and using the \rs{switching}{anisotropic} model. The location of the peak linear growth rate is also shown.}
\label{fig:stability_and_growth}%
\end{figure}

@@

\begin{figure}
  \centering
    \begin{subfigure}{0.49\textwidth}
      \includegraphics[width=\linewidth]{perturbations_4.pdf}
      \caption{Perturbations}
      \label{fig:pressure_pert_4}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.49\textwidth}
      \includegraphics[width=\linewidth]{resonant_surface_4.pdf}
      \caption{Resonance function}
      \label{fig:resonant_surface_4}
    \end{subfigure}
\mycaption{Perturbations corresponding to the flute and
  kink instabilities and the spatial radial distribution of the
  associated resonance function.}{Pressure and velocity perturbations
  in $z$ (corresponding to the flute and kink
  instabilities, respectively) and of the resonance
  \rs{function}{surface} $m B_{\theta}(r)/r + kB_z(r)$ as a function
  of $r$ using the observed fluting perturbation wavenumbers. All
  plots are \rs{from}{} snapshots at $t=26$ where $\eta=10^{-4}$ and
  the viscosity model is \rs{switching}{anisotropic}.} 
\label{fig:k_and_resonance}%
\end{figure}

\rs{Figure~\ref{fig:pressure_pert_4} plots the }{The} observed
  perturbations corresponding to the flute and kink
  instabilities at $t=26$ \rs{}{are shown in
  Figure~\ref{fig:pressure_pert_4}}. The fluting perturbation is
  most easily observed in the pressure and is plotted as a function of $z$
  following a line through the point $(r, \theta) = (0.101, 0)$. The
  kink instability is best revealed in the $x$-velocity (a proxy for the
  radial velocity) through the axis. Comparing the magnitudes of the
  perturbations at this time suggests the flute
  instability is close to transitioning to its nonlinear phase while
  the kink instability is still very much in its linear
  phase.

@@

Comparing the effect of the viscous models on the perturbations, in the isotropic case, the fluting perturbation is damped, while in the \rs{switching}{anisotropic} case the kink perturbation is diminished, explaining why the flute instability appears more readily in the \rs{switching}{anisotropic} case (figure~\ref{fig:kink_ke-4}).
